What Boolean Algebras Can Teach Us About Weighted Model Counting

### 2020-04-12 paper.tex
Free algebras and pushouts/coproducts

\title{Generalising Weighted Model Counting to Boolean Algebras}
%\title{On the Limitations of Weighted Model Counting}
%\title{Weighted Model Counting/Integration from the Perspective of Boolean Algebras}
%\title{What Boolean Algebras Can Teach Us About Weighted Model Counting/Integration?}
\author{Paulius Dilkas}

@@

@@

% Contributions:
% 1. A generalisation of WMC to Boolean algebras. We show how WMC makes
% independence assumptions and how we can make some but not all literals
% independent.
% 1.5. We show how the lifted inference techniques stem from the underlying
% algebraic structure.
% 1.6. Do we add more inference rules?
% 2. We show that it is a valid measure.
% 3. We show that WMC assumes independence, i.e., cannot represent all measures.
% 4. We show how the BA can be extended with new 'literals' to represent any measure.
% 5. (Maybe) we establish a lower bound on the number of new literals, if we
% make no assumptions about independence.
% 6. show how evidence = quotiening by a filter, and how quotiening by an ideal works
% 7. (WMI) If all weights are 1, we extend Boolean variables to discrete
% variables, and continuous variables are arbitrary subsets (in the
% sigma-algebra), then the WMI process calculates the probability of any event
% and doesn't require going down to the level of models (assuming a suitable
% probability measure).
% 8. If the weight of a literal (and its negation) is 1, then we don't need to
% look at its models--the probability can be calculated at any level. This
% extends to any number of variables.
% 9. How WMI transforms to our algebras. The boring version and what's wrong
% with it.
% 10. Show how to transfer the measure to a quotient algebra and that the answer
% match with each other and with the original definition of WMC.

Potential directions to explore (no good ones, really):
\begin{itemize}
\item Infinite BAs.
  % TODO: Feedback 2: understanding this is the key for formulating the
  % relevance/significance of this work
  \begin{itemize}
  \item For example, the BA of finite and cofinite sets could be interesting.
  \item OUWMC requires infinite logics to be compact (whatever that means). My
    algebraic angle suggests that completeness should be enough. Topologically,
    compactness implies completeness, but I have no idea how this translates to
    logics and algebras.
  \end{itemize}
\item WMI
\item Measures with something other than $\mathbb{R}_{\ge 0}$ as the codomain
  (doesn't look promising).
\end{itemize}

Previous/related work:
\begin{itemize}
\item Hailperin's approach to probability logic
  \cite{DBLP:journals/ndjfl/Hailperin84}
\item Nilsson's (somewhat successful) probabilistic logic
  \cite{DBLP:journals/ai/Nilsson86}
\item Logical induction: a big paper with a good overview of previous attempts
  to assign probabilities to logical sentences in a sensible way
  \cite{DBLP:journals/eccc/GarrabrantBCST16}
\item Semiring programming \cite{DBLP:journals/corr/BelleR16}
\item WMI \cite{DBLP:conf/ijcai/BellePB15}
\item Measures on Boolean algebras
  \begin{itemize}
  \item On possibility and probability measures in finite Boolean algebras
    \cite{DBLP:journals/soco/CastineiraCT02}
  \item Representation of conditional probability measures
    \cite{krauss1968representation}
  \end{itemize}
\end{itemize}

@@

\begin{lemma}[\cite{givant2008introduction}] \label{thm:representation}
  The following are equivalent:
  \begin{itemize}
  \item $\mathbf{B}$ is atomic.
  \item For any $x \in \mathbf{B}$,
    \[
      x = \bigvee_{\text{atoms } a \le x} a.
    \]
  \item $1$ is the supremum of all atoms.
  \end{itemize}
\end{lemma}

@@

% TODO: make up my mind about a,b vs x,y and stick to it (maybe x,y?)
% TODO: perhaps reorder things into paragraphs, i.e., a paragraph for order, for
% homomorphisms, for ideals and quotients. This would take up less space.

% TODO: Feedback 2: I did a high level read to get the big picture but found
% this discussion on ideals distracting. Specifically, w.r.t. the main claims
% when do these things get used?

@@

% TODO: Feedback 2: Result #1: What are the conditions under which WMC defines a
% measure as introduced above? [Non-negativity should be enough so this is not a
% result.]

@@

% TODO: Feedback 2: You need to explain what precisely these mean in logic and
% models and weight functions are usually defined and understood.
\todo[inline]{This should be replaced with inner sums (a.k.a. free products)}
\begin{definition} \label{def:wmc}
  Let $\mathbf{B}$ be an atomic BA, and let $M \subset \mathbf{B}$ be its set of
  atoms. Let $L \subset \mathbf{B}$ be such that every atom $m \in M$ can be
  uniquely expressed as $m = \bigwedge L'$ for some $L' \subseteq L$, and let
  $w\colon L \to \mathbb{R}_{\ge 0}$ be arbitrary. The \emph{weighted model
    count} $\WMC_w\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC_w(x) = \begin{cases}
      0 & \text{if } x = 0 \\
      \prod_{l \in L'} w(l) & \text{if } M \ni x = \bigwedge L' \\
      \sum_{\text{atoms } a \le x} \WMC_w(a) & \text{otherwise}
    \end{cases}
  \]
  for any $x \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC_w\colon \mathbf{B} \to [0, 1]$ as $\nWMC_w(x) =
  \frac{\WMC_w(x)}{\WMC_w(1)}$ for all $x \in \mathbf{B}$. For both $\WMC_w$ and
  $\nWMC_w$, we will drop the subscript when doing so results in no potential
  confusion.
%  Finally, we say that a measure $m\colon \mathbf{B} \to
%  \mathbb{R}_{\ge 0}$ is a \emph{WMC measure} (or is \emph{WMC-computable}) if
%  there exists a subset $L \subset \mathbf{B}$ and a weight function $w\colon L
%  \to \mathbb{R}_{\ge 0}$ such that $m = \WMC_w$.
\end{definition}
% TODO: mention that the definition can be reduced to a single formula (i.e.,
% without cases)
% TODO: any measure is a WMC measure if all atoms are in L

@@

% TODO: Feedback 2: I look the significance of the below results. Proving
% properties that don't [???] the "roadmap" means (AI) readers will lose
% interest. [This applies until the beginning of the next section.]

@@

  The two remaining properties are easy to prove:
  \begin{itemize}
  \item If $m(1) = 1$, then $m^*(1/(p)) = m(1 \lor p) = m (1) = 1$.
  \item Suppose that $m$ is strictly positive, and let $a/(p) \in
    \mathbf{B}/(p)$ be such that $a/(p) \ne 0/(p)$. Then
    \[
      m^*(a/(p)) = m(a \lor p) \ge m(a) > 0,
    \]
    where the first inequality comes from an elementary property of $\le$ that
    $x \le x \lor y$ for any $x, y \in \mathbf{B}$ \cite{sikorski1969boolean}
    and \cref{lemma:measure_and_order}; and the second inequality follows
    because $a/(p) \ne 0/(p)$ implies that $a \ne 0$, and $m$ is assumed to be
    strictly positive.
  \end{itemize}
\end{proof}
% TODO: add parentheses around quotients (e.g., (a+b)/I).
% TODO: I think I need to cite something for: a/I = 0/I <=> a \in I

@@

% TODO: start here
%\[
%  \begin{tikzcd}
%    \mathbf{B}/(p) \arrow[rr, shift left, "x/(p) \mapsto x \lor p"] \arrow[rr,
%    shift right, swap, "x/(p) \mapsto x \land \neg p"] & & \mathbf{B} \arrow{r}{\WMC} & \mathbb{R}_{\ge 0}
%  \end{tikzcd}
%\]

@@

% TODO: how can WMC compute something in the quotient algebra? Add an example.
% TODO: we can then move through the quotient BA using representatives we get
% from mapping true things to 1 and false things to 0.

% TODO: Feedback: Can you say something here about factorized vs non-factorized
% weight function definitions? That is, factorized is when w maps literals to
% R_>=0, non-factorized is when w maps models to R_>=0 and show:
% a) come up with nice example when non-factorized weights are intuitive
% b) what if weight functions are negative/complex?
% c) clarify that the factorized definition have is w.r.t. models, in case some
% one gets confused [It doesn't have to be, if the BA is not free -- P.]
% d) can you say something about WMI

@@

\todo[inline]{Proofs need to be updated and propositions could be phrased in a
  better way, but the gist should be the same.}

% TODO: Feedback 2: Result #2: WMC c measures. That is, WMC wight weights on
% literals is less expressive than (a) WMC with weights on models is less
% expressive than (b) measures on Boolean algebras?  Results only link at (a).
% Is (c) true? Proof? [It could cover (some sort of) infinite cases...]

@@

In simpler terms, any measure can be computed using WMC by extending the Boolean
algebra with more literals. More precisely, we are given the red part in
\[
  \begin{tikzcd}
    \textcolor{red}{\mathbb{R}_{\ge 0}} & & \\
    \textcolor{red}{\mathbf{B}} \arrow[red]{u}{m} \arrow{r}{\iota} &
    \mathbf{B'} \arrow{lu}[swap]{m'} & \\
    \textcolor{red}{L} \arrow[Subset,red]{u}{} \arrow[Subset]{r}{} & L'
    \arrow[Subset]{u}{} \arrow{r}{w} & \mathbb{R}_{\ge 0}
  \end{tikzcd}
\]
and construct the black part in such a way that the triangle commutes.

@@

% TODO: Feedback 2: You need to explain the significance of this result. Result
% #3: Augmented approach recovers the expressivity of (b). [This is a result,
% although not a big one.]

@@

\bibliographystyle{plain}
\bibliography{paper}
% TODO: curly brackets or parentheses for families? Maybe curly brackets

>>>>

\title{What Boolean Algebras Can Teach Us About Weighted Model Counting}
%\title{Generalising Weighted Model Counting to Boolean Algebras}
%\title{On the Limitations of Weighted Model Counting}
%\title{Weighted Model Counting/Integration from the Perspective of Boolean Algebras}
%\title{What Boolean Algebras Can Teach Us About Weighted Model Counting/Integration?}
\author{Paulius Dilkas}

@@

% Feedback: Can you say something here about factorized vs non-factorized
% weight function definitions? That is, factorized is when w maps literals to
% R_>=0, non-factorized is when w maps models to R_>=0 and show:
% a) come up with nice example when non-factorized weights are intuitive
% b) what if weight functions are negative/complex?
% c) clarify that the factorized definition have is w.r.t. models, in case some
% one gets confused [It doesn't have to be, if the BA is not free -- P.]

@@

% Feedback 2: I did a high level read to get the big picture but found
% this discussion on ideals distracting. Specifically, w.r.t. the main claims
% when do these things get used?

% TODO:
% 1) theorem: a pushout of free algebras is a free algebra. Same for
% coproducts. This extends to coproducts because they are a special case of
% pushouts.
% TODO for later:
% 2) curly brackets or parentheses for families? Maybe curly brackets
% 3) make up my mind about a,b vs x,y and stick to it (maybe x,y?)
% 4) add parentheses around quotients (e.g., (a+b)/I).
% 5) perhaps reorder the section of preliminaries into paragraphs, i.e., a
% paragraph for order, for homomorphisms, for ideals and quotients. This would
% take up less space.

\paragraph{Contributions.}
\begin{itemize}
\item WMC defines a measure over a BA.
\item WMC with weights on literals imposes an independence assumption. (Measures
  are `slightly' more expressive than WMC with weights on models because they
  apply to non-atomic BAs.)
\item A BA can be augmented with new literals in order to support any measure.
\item (Maybe) a lower bound on the number of new literals needed in order to
  support any measure.
\item Alternatively, one can use coproducts and pushouts to define a BA with
  precisely the right independence and conditional independence conditions.
  (This requires a relaxed version of WMC.)
\item This results in a smaller problem for WMC algorithms (w.r.t. both the
  number of literals and the length of the theory) and is optimal for, e.g.,
  Bayesian networks.
\item (Maybe) this results in faster inference (?)
\end{itemize}

\paragraph{Notable previous/related work.}
\begin{itemize}
\item Hailperin's approach to probability logic
  \cite{DBLP:journals/ndjfl/Hailperin84}
\item Nilsson's (somewhat successful) probabilistic logic
  \cite{DBLP:journals/ai/Nilsson86}
\item Logical induction: a big paper with a good overview of previous attempts
  to assign probabilities to logical sentences in a sensible way
  \cite{DBLP:journals/eccc/GarrabrantBCST16}
\item Measures on Boolean algebras
  \begin{itemize}
  \item On possibility and probability measures in finite Boolean algebras
    \cite{DBLP:journals/soco/CastineiraCT02}
  \item Representation of conditional probability measures
    \cite{krauss1968representation}
  \end{itemize}
\end{itemize}

@@

\begin{lemma}[\cite{givant2008introduction}] \label{thm:representation}
  The following are equivalent:
  \begin{itemize}
  \item $\mathbf{B}$ is atomic.
  \item For any $x \in \mathbf{B}$, $x = \bigvee_{\text{atoms } a \le x} a$.
  \item $1$ is the supremum of all atoms.
  \end{itemize}
\end{lemma}

@@

@@

@@

% TODO: Feedback 2: You need to explain what precisely these mean in logic and
% models and weight functions are usually defined and understood.
% TODO: This should be replaced with inner sums (a.k.a. free products)
\begin{definition} \label{def:wmc}
  Let $\mathbf{B}$ be an atomic BA, and let $M \subset \mathbf{B}$ be its set of
  atoms. Let $L \subset \mathbf{B}$ be such that every atom $m \in M$ can be
  uniquely expressed as $m = \bigwedge L'$ for some $L' \subseteq L$, and let
  $w\colon L \to \mathbb{R}_{\ge 0}$ be arbitrary. The \emph{weighted model
    count} $\WMC_w\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC_w(x) = \begin{cases}
      0 & \text{if } x = 0 \\
      \prod_{l \in L'} w(l) & \text{if } M \ni x = \bigwedge L' \\
      \sum_{\text{atoms } a \le x} \WMC_w(a) & \text{otherwise}
    \end{cases}
  \]
  for any $x \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC_w\colon \mathbf{B} \to [0, 1]$ as $\nWMC_w(x) =
  \frac{\WMC_w(x)}{\WMC_w(1)}$ for all $x \in \mathbf{B}$. For both $\WMC_w$ and
  $\nWMC_w$, we will drop the subscript when doing so results in no potential
  confusion.
%  Finally, we say that a measure $m\colon \mathbf{B} \to
%  \mathbb{R}_{\ge 0}$ is a \emph{WMC measure} (or is \emph{WMC-computable}) if
%  there exists a subset $L \subset \mathbf{B}$ and a weight function $w\colon L
%  \to \mathbb{R}_{\ge 0}$ such that $m = \WMC_w$.
\end{definition}
% TODO: mention that the definition can be reduced to a single formula (i.e.,
% without cases)
% TODO: any measure is a WMC measure if all atoms are in L

@@

% TODO: Feedback 2: I [???] the significance of the below results. Proving
% properties that don't [???] the "roadmap" means (AI) readers will lose
% interest. [This applies until the beginning of the next section.]

@@

  The two remaining properties are easy to prove:
  \begin{itemize}
  \item If $m(1) = 1$, then $m^*(1/(p)) = m(1 \lor p) = m (1) = 1$.
  \item Suppose that $m$ is strictly positive, and let $a/(p) \in
    \mathbf{B}/(p)$ be such that $a/(p) \ne 0/(p)$. Then
    \[
      m^*(a/(p)) = m(a \lor p) \ge m(a) > 0,
    \]
    where the first inequality comes from an elementary property of $\le$ that
    $x \le x \lor y$ for any $x, y \in \mathbf{B}$ \cite{sikorski1969boolean}
    and \cref{lemma:measure_and_order}; and the second inequality follows
    because $a/(p) \ne 0/(p)$ implies that $a \ne 0$, and $m$ is assumed to be
    strictly positive.
  \end{itemize}
\end{proof}
% TODO: I think I need to cite something for: a/I = 0/I <=> a \in I

@@

% we can then move through the quotient BA using representatives we get
% from mapping true things to 1 and false things to 0.

%\[
%  \begin{tikzcd}
%    \mathbf{B}/(p) \arrow[rr, shift left, "x/(p) \mapsto x \lor p"] \arrow[rr,
%    shift right, swap, "x/(p) \mapsto x \land \neg p"] & & \mathbf{B} \arrow{r}{\WMC} & \mathbb{R}_{\ge 0}
%  \end{tikzcd}
%\]

@@

@@

% TODO: Proofs need to be updated and propositions could be phrased in a better
% way, but the gist should be the same.

@@

In simpler terms, any measure can be computed using WMC by extending the Boolean
algebra with more literals. More precisely, we are given the red part in
\[
  \begin{tikzcd}
    \textcolor{red}{\mathbb{R}_{\ge 0}} & & \\
    \textcolor{red}{\mathbf{B}} \arrow[red]{u}{m} \ar[r,hookrightarrow,"\iota"]
    & \mathbf{B'} \arrow{lu}[swap]{m'} & \\
    \textcolor{red}{L} \ar[u,red,hookrightarrow] \ar[r,hookrightarrow] & L'
    \ar[u,hookrightarrow] \arrow{r}{w} & \mathbb{R}_{\ge 0}
  \end{tikzcd}
\]
and construct the black part in such a way that the triangle commutes.

@@

% TODO: Feedback 2: You need to explain the significance of this result.

@@

\begin{theorem}
  The number of weights needed to encode a Bayesian network using
  coproducts and pushouts is equal to the number of entries in the tables of the
  network (and the resulting theory is shorter).
\end{theorem}

% TODO: make sure that the definition of 'free' is sufficient
\begin{theorem}[Pushouts of free BAs are free]
  Let
  \[
    \mathbf{A} = \bigoplus_{\substack{\mathbf{B}\\ i \in I}} \mathbf{A}_i
  \]
  be an amalgamated free product such that $\{ \mathbf{A}_{i} \}_{i \in I}$ are
  free BAs with $\{ S_i \}_{i \in I}$ as their respective sets of generators.
  Let $S = \bigcup_{i \in I} S_i$. Then $\mathbf{A}$ is a free BA with
  generating set $S$.
\end{theorem}
\begin{proof}
  Suppose we have a map from $S$ to an arbitrary BA $\mathbf{C}$, as in
  \[
    \begin{tikzcd}
      S_i \ar[r,hookrightarrow] \ar[d,hookrightarrow] & S \ar[d,hookrightarrow]
      \ar[ddr,bend left] & \\
      \mathbf{A}_i \ar[r,hookrightarrow] \ar[rrd,dashed,bend right] & \mathbf{A}
      \ar[rd,dashed]& \\
      & & \mathbf{C}.
    \end{tikzcd}
  \]
  We want to show that there exists a unique homomorphism $\mathbf{A} \to
  \mathbf{C}$. For $i \in I$, from $S_i \hookrightarrow S$ and $S \to
  \mathbf{C}$ we get a map $S_i \to \mathbf{C}$, so---by the definition of a
  free BA---there is a unique homomorphism $\mathbf{A}_i \to \mathbf{C}$.
  Furthermore, a family of homomorphisms $\{ \mathbf{A}_i \to \mathbf{C} \}_{i
    \in I}$ uniquely determine a homomorphism $\mathbf{A} \to \mathbf{C}$ by the
  universal mapping property of a (wide) pushout. Thus $\mathbf{A}$ is a free BA
  with generating set $S$.
\end{proof}

\begin{corollary}
  Similarly, coproducts of free BAs are free.
\end{corollary}

\bibliographystyle{plain}
\bibliography{paper}
